Follow-up Samples

### 2017-04-17 submission/manuscript.Rmd
Editing

***The effects of treatment are not consistent across treatment groups.***
We used two approaches to identify those bacterial populations that change between the two samples for each diagnosis group. First, we sought to identify individual OTUs that could account for the change in overall community structure. However, using a paired Wilcoxon test we were unable to identify any OTUs that were significantly different in the pre and post-treatment groups [Figure S1]. It is likely that high inter-individual variation and the patchy distribution of OTUs across individuals limited the statistical power of the test. To overcome these problems we developed Random Forest models to identify collections of OTUs that would allow us to differentiate between pre and post-treatment samples from each of the diagnosis groups. Because of the relatively small number of subject in each group, we restricted our models to only incorporate 10 OTUs. Despite this restriction, the models performed well (adenoma AUC=`r format(min(adn_treat_AUC_summary$ROC), digits = 2)` - `r format(max(adn_treat_AUC_summary$ROC), digits = 2)`, advanced adenoma AUC=`r format(min(srn_treat_AUC_summary$ROC), digits = 2)` - `r format(max(srn_treat_AUC_summary$ROC), digits = 2)`, carcinoma AUC=`r format(min(crc_treat_AUC_summary$ROC), digits = 2)` - `r format(max(crc_treat_AUC_summary$ROC), digits = 2)`). Interestingly, the 10 OTUs that were used for each model had little overlap with each other [Figure 2]. These results support the earlier community-wide analysis where we observed that the treatment had an impact on the overall community structure; however, the effect of treatment was not consistent across patients and diagnosis groups.

***Post-treatment samples from patients with carcinoma share***
Next, we determined whether treatment changed the microbiota in a way that the post-treatment communities resembled that of patients with normal colons. To test this, we used an expanded cohort of 423 individuals that were diagnosed under the same protocol as having normal colons or colons with adenoma, advanced adenoma, or carcinoma (Table 2). We then constructed Random Forest models to classify the pre and post-treatment samples as having their original diagnosis or having a normal colon. The models performed well (adenoma AUC=`r format(min(adn_AUC_run_summary$ROC), digits = 2)` - `r format(max(adn_AUC_run_summary$ROC), digits = 2)`, advanced adenoma AUC=`r format(min(srn_AUC_run_summary$ROC), digits = 2)` - `r format(max(srn_AUC_run_summary$ROC), digits = 2)`, carcinoma AUC=`r format(min(crc_AUC_run_summary$ROC), digits = 2)` - `r format(max(crc_AUC_run_summary$ROC), digits = 2)`; Figure S2). The OTUs that were most important in the models for differentiating between patients with normal colons and those with adenomas or advanced adenomas were dominated by populations commonly considered as commensals (e.g. *Faecalibacterium*, *Lachnospiraceae*, *Bacteroides*, *Dorea*, *Anaerostipes*, and *Roseburia*) [Figure S3]. Although many of these OTUs were also included in the model differentiating between patients with normal colons and those with carcinoma, this model also included OTUs affiliated with populations that have previously been associated with carcinoma (*Fusobacterium*, *Porphyromonas*, *Parvimonas*) [Figure S3]. Finally, we applied these three models to the pre and post-treatment samples for each diagnosis group and quantified the change in the positive probability of the model. A decrease in the positive probability would indicate that the colon more closely resembled that of a patient with a normal colon. There was not a significant change in the positive probability for the adenoma or advanced adenoma groups [Figure 3]. The positive probability for the pre and post-treatment samples from patients diagnosed with carcinoma significantly decreased. In fact, only `r table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` of the 26 patients (`r round(table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]/26*100,digits = 2)`%) that were initially diagnosed with a carcinoma had a higher positive probability after treatment. One of those was re-diagnosed with carcinoma on the follow up visit. These results indicate that although there are changes in the microbiota associated with treatment, those experienced by the patients in the carcinoma group were the only ones where the change was directed towards what would be found in a normal colon

@@

## Discussion
This study builds upon previous work from numerous labs that have considered both how the bacterial community between those with and without CRC differs and how it might be used as an early screening tool [@yu_metagenomic_2017; @baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014; @warren_co-occurrence_2013]. Here we show that the microbiome changes towards normal after treatment for carcinoma and that chemotherapy and radiation did not provide any additional change. Although some of the important OTUs classified to genera from bacteria considered the "usual suspects" (e.g. *Fusobacterium*, *Porphyromonas*, and *Parvimonas*) many did not. The majority of important OTUs had taxonomic classifications for resident gut microbes and were common for the adenoma, advanced adenoma, and carcinoma models. This suggests that members within the commensal community may be the first that change during CRC pathogenesis. These subtle changes, in turn, could be the first step in allowing more inflammatory bacteria to gain a foothold within the colon [@flynn_metabolic_2016].

Unlike previous studies on the microbiome and CRC, ours focuses on comparing the microbiome during recovery from treatment in adenoma, advanced adenoma, and carcinoma groups. Although there were differences for genera associated with specific bacterium linked with CRC [Figure S3], most of the important OTUs taxonomically classified to commensal bacteria [Figure S3]. Although these changes may be subtle, due to the lack of significant differences in the bacterial community pre- and post-treatment in adenoma and advanced adenoma [Figure 1]; they support the hypothesis that the first members of the community to change and potentially stay changed even after treatment are those that are commensal bacteria.

@@

A limitation in our study was that there was a significant difference in the time elapsed in the collection of the post-treatment sample between adenoma or advanced adenoma versus carcinoma (P-value `r ifelse(time_pvalues["adnVcrc", "bh"] < 0.05 & time_pvalues["srnVcrc", "bh"] < 0.05, invisible("<"), invisible(">"))` 0.05), with time passed being less for adenoma (`r paste(format(time_summary["adenoma", "mean_days"], digits = 0), label = "\u00B1", format(time_summary["adenoma", "sd_days"], digits = 0))` days) and advanced adenoma (`r paste(format(time_summary["srn", "mean_days"], digits = 0), label = "\u00B1", format(time_summary["srn", "sd_days"], digits = 0))`) than carcinoma (`r paste(format(time_summary["carcinoma", "mean_days"], digits = 0), label = "\u00B1", format(time_summary["carcinoma", "sd_days"], digits = 0))`). These results would indicate that some of the differences observed between the carcinoma and adenoma groups could be due to differences in collection time. Specifically, it could confound the observation that carcinomas changed more than adenomas [Figure 1]. However, there are two reasons that this may not be the case. First, the advanced adenoma group had a higher dissimilarity than adenoma but lower dissimilarity then carcinoma and the collection time to their follow up samples was less than the adenoma group. Second, this confounding would not affect the observations where models were used since they were built using a different cohort [Figures 2 & S2-S3].

Another limitation of this study was that it drew heavily from those with Caucasian ancestry making it possible that the observations may not be representative of those with either Asian or African ancestry. Although our training and test set are relatively large we still run the risk of over-fitting or having a model that may not be representative of other populations. We have done our best to safeguard against this by not only running 10-fold cross validation but also having over 100 different 80/20 splits to try and mimic the type of variation that might be expected to occur.

>>>>

***The effects of treatment are not consistent across treatment groups.***
We used two approaches to identify those bacterial populations that change between the two samples for each diagnosis group. First, we sought to identify individual OTUs that could account for the change in overall community structure. However, using a paired Wilcoxon test we were unable to identify any OTUs that were significantly different in the pre and post-treatment groups [Figure S1]. It is likely that high inter-individual variation and the patchy distribution of OTUs across individuals limited the statistical power of the test. To overcome these problems we developed Random Forest models to identify collections of OTUs that would allow us to differentiate between pre and post-treatment samples from each of the diagnosis groups. Because of the relatively small number of subject in each group, we restricted our models to only incorporate 10 OTUs to limit the likelihood that that models would overfit the data. Despite this restriction, the models performed well (adenoma AUC=`r format(min(adn_treat_AUC_summary$ROC), digits = 2)` - `r format(max(adn_treat_AUC_summary$ROC), digits = 2)`, advanced adenoma AUC=`r format(min(srn_treat_AUC_summary$ROC), digits = 2)` - `r format(max(srn_treat_AUC_summary$ROC), digits = 2)`, carcinoma AUC=`r format(min(crc_treat_AUC_summary$ROC), digits = 2)` - `r format(max(crc_treat_AUC_summary$ROC), digits = 2)`). Interestingly, the 10 OTUs that were used for each model had little overlap with each other [Figure 2]. These results support the earlier community-wide analysis where we observed that the treatment had an impact on the overall community structure; however, the effect of treatment was not consistent across patients and diagnosis groups.

***Post-treatment samples from patients with carcinoma share***
Next, we determined whether treatment changed the microbiota in a way that the post-treatment communities resembled that of patients with normal colons. To test this, we used an expanded cohort of 423 individuals that were diagnosed under the same protocol as having normal colons or colons with adenoma, advanced adenoma, or carcinoma (Table 2). We then constructed Random Forest models to classify the pre and post-treatment samples as having their original diagnosis or having a normal colon. The models performed well (adenoma AUC=`r format(min(adn_AUC_run_summary$ROC), digits = 2)` - `r format(max(adn_AUC_run_summary$ROC), digits = 2)`, advanced adenoma AUC=`r format(min(srn_AUC_run_summary$ROC), digits = 2)` - `r format(max(srn_AUC_run_summary$ROC), digits = 2)`, carcinoma AUC=`r format(min(crc_AUC_run_summary$ROC), digits = 2)` - `r format(max(crc_AUC_run_summary$ROC), digits = 2)`; Figure S2). The OTUs that were incorporated into the adenoma and advanced adenoma models largely overlapped and those OTUs that were used to classify the carcinoma samples were largely distinct from those of the other two models [Figure 4A]. Among the OTUs that were shared across the three models were those populations commonly considered as commensals (e.g. *Faecalibacterium*, *Lachnospiraceae*, *Bacteroides*, *Dorea*, *Anaerostipes*, and *Roseburia*) [Figures 4B]. Although many of these OTUs were also included in the model differentiating between patients with normal colons and those with carcinoma, this model also included OTUs affiliated with populations that have previously been associated with carcinoma (*Fusobacterium*, *Porphyromonas*, *Parvimonas*) [Figure S3]. Finally, we applied these three models to the pre and post-treatment samples for each diagnosis group and quantified the change in the positive probability of the model. A decrease in the positive probability would indicate that the colon more closely resembled that of a patient with a normal colon. There was not a significant change in the positive probability for the adenoma or advanced adenoma groups [Figure 3]. The positive probability for the pre and post-treatment samples from patients diagnosed with carcinoma significantly decreased. In fact, only `r table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` of the 26 patients (`r round(table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]/26*100,digits = 2)`%) that were initially diagnosed with a carcinoma had a higher positive probability after treatment. One of those was re-diagnosed with carcinoma on the follow up visit. These results indicate that although there are changes in the microbiota associated with treatment, those experienced by the patients in the carcinoma group were the only ones where the change was directed towards what would be found in a normal colon

@@

## Discussion

This study builds upon previous work that demonstrated a role for the microbiota in tumorigenesis and that the microbiome contains biomarkers to diagnosis colonic lesions [ADD REFERENCES FOR CAUSATION;  @yu_metagenomic_2017; @baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014; @warren_co-occurrence_2013]. Our study focused on comparing the microbiome of patients diagnosed with adenoma, advanced adenoma, and carcinoma before and after treatment. For all three groups of patients we observed changes in their microbiota. The microbiota of patients diagnosed with carcinoma changed significant more than the other groups. This change resulted in communities that more closely resembled those of patients with normal colons.

* Does treatment affect the colonic microbiota in a predictable manner? For those with carcinoma

* Does the treatment alter the community to more closely resemble that of individuals with normal colons? For those with carcinoma.

* Heterogeneity of the population (common problem for all microbiome studies) made it impossible to identify speicfi microbiota that changed with treatment; however, RF made it possible to identify consortia of microbiota that were associated with treatment and recovery.

* End.

Although there were differences for genera associated with specific bacterium linked with CRC [Figure S3], most of the important OTUs taxonomically classified to commensal bacteria [Figure S3]. Although these changes may be subtle, due to the lack of significant differences in the bacterial community pre- and post-treatment in adenoma and advanced adenoma [Figure 1]; they support the hypothesis that the first members of the community to change and potentially stay changed even after treatment are those that are commensal bacteria.

Here we show that the microbiome changes towards normal after treatment for carcinoma

Although some of the important OTUs classified to genera from bacteria considered the "usual suspects" (e.g. *Fusobacterium*, *Porphyromonas*, and *Parvimonas*) many did not. The majority of important OTUs had taxonomic classifications for resident gut microbes and were common for the adenoma, advanced adenoma, and carcinoma models. This suggests that members within the commensal community may be the first that change during CRC pathogenesis. These subtle changes, in turn, could be the first step in allowing more inflammatory bacteria to gain a foothold within the colon [@flynn_metabolic_2016].

@@
